---
title: "Exploring U.S. Supreme Court Decisions"
author: "Dylan Petiprin"
output: github_document
---

## Get the data

```{r}
# load useful packages
library(tidyverse)

# load data
setwd("~/CFSS Homework/hw03/hw03/data")
vote <- read_csv("scdb-vote.csv")
case <- read_csv("scdb-case.csv")
```


## What percentage of cases in each term are decided by a one-vote margin (i.e. 5-4, 4-3, etc.)


```{r}
# Calculate the number of 1-vote margins per term

# create a list of one-vote proportions using summarise()
case %>% 
  group_by(term) %>% 
  select(c(term,majVotes,minVotes)) %>% 
  drop_na(minVotes) %>% 
  mutate(vote_diff = majVotes - minVotes) %>% 
  summarise(one_vote_perc = sum(vote_diff==1)/sum(term=n())) %>% 
  
# put into geom_line()
  ggplot(aes(term, one_vote_perc))+
  geom_line()+
  labs(title="Percent of U.S. Supreme Court cases decided by 1-vote margin",
       x="Term", y="Percent of total cases decided")

```

## For justices [currently serving on the Supreme Court](https://www.supremecourt.gov/about/biographies.aspx), how often have they voted in the conservative direction in cases involving criminal procedure, civil rights, economic activity, and federal taxation?

Organize the resulting graph by justice in descending order of seniority. (Note that the chief justice is always considered the most senior member of the court, regardless of appointment date.)

```{r}

#c alculate proportion of conservative votes by issue area
vote %>% 
    left_join(case, by = "caseId") %>% 
    select(justiceName, direction, issueArea) %>% 
  filter(justiceName %in% c("BMKavanaugh","NMGorsuch","EKagan","SSotomayor","SAAlito","SGBreyer","CThomas","JGRoberts")) %>% 
  filter(issueArea %in% c(1,2,8,12)) %>% 
    drop_na(direction) %>% 
  group_by(justiceName, issueArea) %>% 
  summarise(cons_perc = sum(direction == 1)/sum(justiceName = n())) %>% 
  
# relabel issue areas in English and place justices by seniority
  mutate(issueArea = factor(issueArea, levels = c(12,8,2,1), 
       labels =c("Federal Taxation","Economic Activity","Civil Rights","Criminal Procedure"))) %>% 
  mutate(justiceName = factor(justiceName, levels = c("JGRoberts","CThomas","SGBreyer","SAAlito","SSotomayor","EKagan","NMGorsuch","BMKavanaugh"))) %>% 
  
# plot using geom_col() and facet by justice
ggplot(aes(issueArea, cons_perc*100))+
  geom_col()+
  coord_flip()+
  facet_wrap(~justiceName)+
  labs(title= "U.S Supreme Court", subtitle="Percent of cases decided in a conservative direction", x="Issue Area",y="Percent of votes cast")


 
  
```

## In each term, how many of the term's published decisions (decided after oral arguments) were announced in a given month?

```{r}

# get decision totals by month, making sure to filter by post-oral arguments
case %>% 
      select(caseIssuesId,dateDecision, term, decisionType) %>% 
  filter(decisionType %in% c(1,6,7)) %>% 
  separate(dateDecision, into = c("Month","Day","Year")) %>% 
  mutate(Month = factor(Month, levels = c(9,8,7,6,5,4,3,2,1,12,11,10),labels = c("September","August","July","June","May","April","March","February","January","December","November","October"))) %>% 
  group_by(Month, term) %>%
 tally() %>% 
  
# plot decision totals using geom_boxplot()
   ggplot(aes(Month,n))+
  geom_boxplot()+
  coord_flip()+
  labs(title = "U.S. Supreme Court", subtitle = "Number of decisions announced post-oral arguments per month, by term",
       y="Number of decisions announced in a term-month")
```

## Which justices are most likely to agree with with the Court's declaration that an act of Congress, a state or territorial law, or a municipal ordinance is unconstitutional? Identify all cases where the Court declared something unconstitutional and determine the ten justices who most and least frequently agreed with this outcome as a percentage of all votes cast by the justice in these cases. Exclude any justice with fewer than 30 votes in cases where the Court's outcome declares something unconstitutional.

```{r}
# get all the instances of declared unconstitutionality and tally up each justice's votes
vote %>% 
  left_join(case, by = "caseIssuesId") %>% 
  select(declarationUncon, caseIssuesId, justiceName, majority) %>% 
  filter(declarationUncon!=1) %>% 
   group_by(justiceName, majority) %>% 
  drop_na(majority) %>%
  summarise(justice_total = n()) %>%
  pivot_wider(names_from = majority, values_from = justice_total) %>% 
  rename(dissent = 2, majority = 3) %>% 
  
# calculate the percentage each justice agreed for unconstitutionality, filter out those with less than 30 total votes, and get the top and bottom 10
  mutate(dissent=replace_na(dissent,0), 
         justice_total = dissent + majority,
         justice_perc = majority / justice_total) %>% 
  filter(justice_total>30) %>% 
  arrange(justice_perc) %>% 
  ungroup() %>% 
  slice(1:10,69:78) %>% 
  
# plot using geom_col()and color by top/bottom distinction
  ggplot(aes(x = reorder(justiceName, justice_perc),y = justice_perc*100))+
  geom_col(stat = "identity", aes(fill = justice_perc*100>75))+
  theme(legend.position = "none")+
  coord_flip()+
  labs(y="Percent of votes agreeing with declaration",
       x="Justice Name",
       title = "U.S Supreme Court",
       subtitle = "Justices most and least frequently in agreement regarding unconstitutionality")
 
```

## In each term he served on the Court, in what percentage of cases was Justice Antonin Scalia in the majority?

```{r}
# calculate the percentage of Scalia in majority by term
vote %>% 
  left_join(case, by = "caseIssuesId") %>% 
  select(term.x, caseIssuesId, majority, justiceName, vote) %>% 
  filter(justiceName=="AScalia") %>% 
  drop_na(majority) %>% 
  group_by(term.x) %>% 

# plot using geom_line()
summarise(AScal_maj_perc = sum(majority == 2)/sum(majority= n())) %>% 
  ggplot(aes(x=term.x, y=AScal_maj_perc*100))+
  geom_line()+
  labs(x = "Term", y = "Percent of total decisions", title= "Percent of the time Scalia was in the majority" )

```

## Create a graph similar to above that adds a second component which compares the percentage for all cases versus non-unanimous cases (i.e. there was at least one dissenting vote)

```{r}
# add another summarise() line to get the percentage for non-unanimous decisions
vote %>% 
  left_join(case, by = "caseIssuesId") %>% 
  select(term.x, caseIssuesId, majority, justiceName, vote, minVotes) %>% 
  filter(justiceName=="AScalia") %>% 
  drop_na(majority) %>% 
  group_by(term.x) %>% 
summarise(AScal_maj_perc = sum(majority == 2)/sum(majority= n()),
          AScal_nunan_maj_perc = sum(majority == 2 & minVotes != 0)/sum(minVotes != 0) ) %>% 

# plot as before, distinguishing lines by color
  ggplot()+
  geom_line(aes(x=term.x, y=AScal_maj_perc*100, color = "All decisions"))+
              geom_line(aes(x=term.x, y=AScal_nunan_maj_perc*100, color = "Non-unanimous decisions")) +
  labs(x = "Term", y = "Percent of total decisions", title= "Percent of the time Scalia was in the majority", color= NULL )+
  theme(legend.position = "bottom")
```

## In each term, what percentage of cases were decided in the conservative direction?

```{r}
# calculate percentage of conservative cases by term
vote %>% 
  left_join(case, by = "caseIssuesId") %>% 
  select(term.x, caseIssuesId, decisionDirection) %>% 
  group_by(term.x) %>% 
  drop_na(decisionDirection) %>% 
  summarise(cons_perc = sum(decisionDirection == 1)/sum(decisionDirection = n())) %>% 
  
# plot using geom_line()
  ggplot(aes(x=term.x, y=cons_perc*100))+
  geom_line()+
  labs(x = "Term", y = "Percent of total decisions", title= "U.S. Supreme Court", subtitle = "Percent of cases decided in the conservative direction" )
```

## The Chief Justice is frequently seen as capable of influencing the ideological direction of the Court. Create a graph similar to the one above that also incorporates information on who was the Chief Justice during the term.

```{r}
# retroactively create new dataframe without chief so that the later plots have that sweet, sweet gray background
vote_nochief <- vote %>% 
  left_join(case, by = "caseIssuesId") %>% 
  select(term.x, decisionDirection, caseIssuesId) %>% 
  group_by(term.x) %>% 
  drop_na(decisionDirection) %>% 
  summarise(cons_perc = sum(decisionDirection == 1)/sum(decisionDirection = n()))

# calculate percentage of conservative votes per chief tenure, and reorder by chief term
 vote %>% 
  left_join(case, by = "caseIssuesId") %>% 
  select(term.x, caseIssuesId, decisionDirection, chief) %>% 
  group_by(term.x, chief) %>% 
  drop_na(decisionDirection) %>% 
  summarise(cons_perc = sum(decisionDirection == 1)/sum(decisionDirection = n())) %>% 
  mutate(chief = fct_reorder(chief, term.x)) %>% 
   
# plot using geom_line(), faceted by chief
  ggplot()+
  geom_line(aes(x=term.x, y=cons_perc*100)) +
  facet_wrap(~chief) +

# add the background layer
   geom_line(data = vote_nochief, aes(x=term.x, y=cons_perc*100), alpha = .25) +
  labs(x = "Term", y = "Percent of total decisions", title= "U.S. Supreme Court", subtitle = "Percent of cases decided in the conservative direction" ) +
  scale_x_continuous(breaks = c(1800,1900,2000))

```



